```{r methods-setup, include=FALSE}
library(knitr)
library(flextable)
```

# 2. Methods

## 2.1 Animals, Housing, and Data Collection

Data were collected as part of a previous study [@sheng2024redefining] using 159 lactating Holstein dairy cows (12 primiparous, 147 multiparous; parity = 3.0 ± 1.1; mean ± SD) enrolled over a 10-month period (July 2020 to May 2021) at The University of British Columbia (UBC) Dairy Education and Research Centre in Agassiz, BC, Canada. All procedures were approved by the UBC Animal Ethics Committee (protocol A19-0299). Cows were housed in a dynamic group of 48 in a pen equipped with 48 sand-bedded lying stalls, 30 electronic feed bins, and 5 electronic water bins (Insentec, Hokofarm, Emmeloord, the Netherlands). The electronic bins identified individual cows via radio frequency identification and recorded the cow identification, bin number, visit start and end times, and start and end weight of the feed or water in the bin for each visit [@chapinal2007validation]. Cows were milked twice daily and total mixed ration was delivered twice daily while cows were away for milking. On average, every 16 ± 3 d, 6 ± 2 cows were removed from the pen and replaced with new animals (i.e., regrouping event); individual cows remained in the trial for an average of 86 ± 54 d. Farm staff performed daily health checks, and a trained observer scored locomotion weekly using a 5-level scale [@flower2006effect] to identify lame cows.

## 2.2 Data Cleaning and Filtering

Data preparation and analysis were conducted using R [version 4.3.1; @R-base]. To ensure that the data we analyzed reflected the baseline patterns of cows when they were healthy, not lame and not in estrus, we applied strict data cleaning and filtering criteria that extended beyond the data cleaning procedure described in Sheng et al. [-@sheng2024redefining]. Starting from 14,728 cow-days across 159 cows and 306 d, exclusions were applied at two levels. At the individual level, data were removed for specific cows on days affected by:

(1) health events, including illness episodes when cows were moved to the sick pen and when cows received veterinary treatment while remaining in the pen (e.g., mastitis, fever, injury), with a buffer period around each event (7 days before and 7 days after the recorded events) to account for potential behavioral changes during the prodromal and recovery periods (16 cows across 120 days, 197 cow-days).

(2) lameness, defined as two consecutive weekly locomotion scores >= 3 [@flower2006effect; @eriksson2020effects], with all days during the lame period removed until the cow was scored as sound again, plus a 7-d buffer before lameness onset (48 cows across 227 days, 679 cow-days).

(3) the first and last day each cow was present in the group, as data on these days would be incomplete (159 cows across 43 days, 285 cow-days).

(4) days when farm staff recorded interference with individual cows due to loss of ear tags, hoof trimming or other routine husbandry procedures, as these events may temporarily alter behavior through handling stress or cows being away from the electronic feed and water bins for extended period of times (9 cows across 22 days, 24 cow-days).

(5) days when cows were not confirmed pregnant, as reproductive status (e.g., bred, fresh, not pregnant) may influence behavior through estrous cycling and associated hormonal fluctuations; only cow-days where the reproductive status was recorded as pregnant were retained (87 cows across 207 days, 2,695 cow-days removed).

At the herd level, data were removed for all cows on days affected by:

(1) estrus events, as any cow in heat can potentially disturb the behavior of all other cows in the group (55 estrus events, 2,381 cow-days) [@reith2018behavioral].

(2) technical disruptions to the electronic feeders and drinkers, including equipment breakdowns, compressor failures, power outages, missing data, and severe weather events that compromised bin function (28 d, 1,281 cow-days).

(3) regrouping events when the herd dynamics was destabilized (16 events, 674 cow-days) [@von2008acute].

After filtering, 107 cows contributed data across 207 days (10-130 days of observation per cow), resulting in 6226 unique cow-days retained.

## 2.3 Behavioral Variable Extraction

To process the raw feeding and drinking data and extract behavioral variables, we developed an open-source R package, 'moo4feed' (<https://www.skysheng.io/moo4feed/>), providing a standardized, customizable workflow for handling electronic feed and water bin data. The package supports five areas of analysis used in this study: (1) data processing, configuration, and quality control; (2) agonistic replacement detection; (3) non-nutritive visit detection; (4) feed availability estimation; and (5) meal clustering. All thresholds and processing parameters described below are user-configurable to cater to different farm setups; the values reported here are those used in the present study. For full documentation, including additional functionalities not used here (e.g., pairwise synchronicity analysis), please refer to the package website.


**Data Processing, Configuration and Quality Control.** Raw visit data were converted into a standardized format, with each record capturing the cow's identity, the bin number, visit start and end times, and the start and end weights of feed or water in the bin. Only days for which both feeder and drinker data files were available were retained. Daylight saving time transitions were detected and corrected automatically. During fall-back transitions (when clocks move back one hour, causing chronological order of visits within the repeated hour to be ambiguous), we removed all visits from the duplicated hour, since it is not possible to determine the true chronological order, and adjusted subsequent timestamps accordingly. During spring-forward transitions (when clocks move forward by one hour), visit timestamps were shifted forward by one hour for the remainder of the day. Records were then passed through a sequence of automated quality control checks, summarized in Table 1.

```{=latex}
\newpage
```

**Table 1.** Automated quality control checks applied to raw visit data from electronic feed and water bins. Dashes (—) indicate that threshold ID is not applicable or is data-driven, and thus no hard-coded numerical threshold applies. Action: Flag = abnormal record is retained and sent to researchers on the farm to review for bin health check; Correct = the record is automatically corrected; Remove = the record is deleted from the dataset. For negative weight corrections, feed or water intake is recalculated from the corrected weights (start weight − end weight); if the recalculated intake remains negative, the record is removed.

```{r echo=FALSE}

qc_tbl <- tibble::tribble(
  ~Check,                            ~Description,                                                                                                                         ~`Feed threshold`, ~`Water threshold`, ~Action,
  "Expected cow count",              "less than 48 cows (expcted cow count) detected, indicating potential lost ear tags for entire day",                                                               "\u2014",          "\u2014",           "Flag",
  "Double detections",               "Same cow recorded at two bins simultaneously, or multiple cows recorded at the same bin simultaneously, indicating potential bin malfunction", "\u2014",   "\u2014",           "Flag and correct",
  "Negative duration or intake",     "Visit records with negative duration or intake values",                                                                               "\u2014",          "\u2014",           "Flag and remove",
  "Negative weight",                 "Negative start or end weight measurements in the bin",                                                                               "\u2014",          "\u2014",           "Correct to 0",
  "Large single-visit intake",       "Abnormal large intake per visit",                                                                                              ">8 kg",           ">30 L",            "Flag",
  "No afternoon feeding or drinking","Cow did not feed or drink after noon, potentially due to a lost ear tag or illness",                                                  "\u2014",          "\u2014",           "Flag",
  "Low or no bin traffic",           "Bins with fewer than 10 visits that day (no visits or unusually low traffic), indicating potential bin malfunction",                "<10 visits/bin/d", "<10 visits/bin/d",  "Flag",
  "Long visit duration",             "Abnormal long visit duration, indicating potential bin malfunction",                                                                                                     ">2,000 s",        ">1,800 s",         "Flag",
  "High intake rate",                "Excessively high intake rate per visit",                                                                                              ">0.008 kg/s",     ">0.35 L/s",        "Flag",
  "Low daily visit count",           "Fewer than 10 visits recorded, suggesting a potentially incomplete day due to bin malfunction",                                                              "<10 visits",      "<10 visits",       "Flag",
  "Abnormal daily intake",           "Abnormally low or high total daily intake",                                                                                           "<35 or >75 kg/d", "<60 or >180 L/d",  "Flag"
)

flextable(qc_tbl) |>
  set_header_labels(
    Check = "Check",
    Description = "Description",
    `Feed threshold` = "Feed threshold",
    `Water threshold` = "Water threshold",
    Action = "Action"
  ) |>
  theme_box() |>
  align(j = c("Check", "Feed threshold", "Water threshold", "Action"),
        align = "center", part = "body") |>
  align(j = "Description", align = "left", part = "body") |>
  align(align = "center", part = "header") |>
  bg(bg = "#6E8B3D", part = "header") |>
  color(color = "white", part = "header") |>
  bold(part = "header") |>
  fontsize(size = 9, part = "all") |>
  width(j = "Check",             width = 1.2) |>
  width(j = "Description",       width = 2.6) |>
  width(j = "Feed threshold",    width = 0.8) |>
  width(j = "Water threshold",   width = 0.8) |>
  width(j = "Action",            width = 0.8) |>
  set_table_properties(layout = "fixed", width = 1)
```


Flagged records were compiled into a daily report and sent to the researchers and students working on the farm at the time of data collection, alerting them to inspect the electronic bins and record any abnormal events (manually noted events listed in Section 2.2). Following these checks, multivariate outlier detection was performed using a k-nearest neighbor (KNN) algorithm [@malini2017analysis] applied to scaled visit duration, intake, and intake rate. Unlike rectangular cutoffs based on standard deviations (SD) of each variable independently, the KNN approach operates in multidimensional space and flags visits that are unusual in combination even when plausible on any single dimension. In this study, k = 50, and the flagging thresholds were set such that 0.04% of feed visits and 0.1% of water visits were identified as outliers.

After cleaning, feed and water datasets were combined, and six conventional daily behavioral variables were calculated per cow: (i) daily feed intake (kg), (ii) daily feeding duration (s), (iii) daily feeding visits, (iv) daily water intake (L), (v) daily drinking duration (s), and (vi) daily drinking visits.

**Agonistic Replacement Detection.** Replacements at the feed bins were identified using a validated algorithm following the method described by Huzzey et al. [-@huzzey2014automatic]. In brief, a replacement was recorded when one cow (the actor) entered a feed bin within 26 seconds of another cow (the reactor) leaving that same bin, serving as a proxy for the actor physically pushing the reactor away from the bin. Replacement events were discarded if the presumed actor was already feeding at a different bin at the time the reactor exited, as this would indicate that the reactor likely left voluntarily rather than being displaced [@foris2019automatic]. Daily agonistic behavior was summarized into two variables: (vii) daily number of times the focal cow acted as the actor (displacing another cow from a bin), and (viii) daily number of times the focal cow acted as the reactor (being displaced from a bin).

**Non-nutritive Visit Detection.** A visit was classified as non-nutritive if the cow's measured intake did not exceed the calibration error of the electronic feeders and drinkers (0.5 kg), meaning the cow visited a bin that contained feed but did not eat anything. This yielded one daily-level variable: (ix) daily non-nutritive/ feed visits per cow.

**Feed Availability Estimation.** Feed delivery events were automatically identified when at least three bins showed a start-weight increase of ≥5 kg within a one-hour window (deliveries on our farm were never longer than 1h). The 5 kg threshold filtered out small spurious weight increases from occasional bin malfunctions or human disturbance (e.g., due to someone leaning on a bin), and requiring at least three bins to increase concurrently to reduce false positives, since these artefacts rarely affect multiple bins at once. All detected delivery times were manually verified by the researchers to confirm accuracy. This allowed us to summarize the frequency of feed deliveries throughout the day and the amount of feed added to each bin during each delivery. By combining delivery records with visit-level bin weight data, we then estimated how much feed remained in each bin at the moment a cow arrived. This produced one variable: (x) median proportion of feed remaining in the bin at visit start (%), calculated as the daily median across all of a cow's feed bin visits. This variable indicates whether a cow consistently arrived at bins with ample fresh feed or tended to access nearly depleted bins, helping flag individuals that may be socially disadvantaged and forced to eat leftover feed.

**Meal Clustering and Analysis.** Individual feed bin visits were grouped into meals using a density-based clustering algorithm, DBSCAN [@khan2014dbscan], which identifies clusters of temporally close visits based on two parameters: maximum time gap between consecutive visits within a meal (the meal criterion), and a minimum number of visits required to form a meal.

To derive the meal criterion, we first extracted all inter-visit intervals (i.e., the time gaps between consecutive feed bin visits from the same animal) pooled across all animals and days. Because these gap times are highly right-skewed (most gaps are short, but a few are very long), a log transformation was applied before fitting a two-component Gaussian Mixture Model (GMM) [@tolkamp1998satiety]. The two components naturally captured two distinct populations of gaps: short within-meal pauses and longer between-meal gaps. The meal criterion was then set to the point where the two fitted distributions intersect. In this study, a single meal criterion was derived based on the inter-visit intervals distribution for all animals and days; the package supports finer granularity (animal-specific or animal-day-specific meal thresholds) and alternative gap-distribution methods, but these were not used here. Meals were required to contain a minimum of two visits. The DBSCAN algorithm then assigned each visit to a meal cluster or flagged it as an isolated, unclustered visit. Timeline visualizations of the resulting meal clusters were generated and visually inspected for each animal-day to confirm clustering quality (example: <https://www.skysheng.io/moo4feed/articles/meal_clustering.html>).

Once visits were grouped into meals, meal-level variables were computed by aggregating visit-level data within each meal and then summarizing across all meals within each cow-day. This step yielded six meal-level variables: (xi) daily meals, (xii) median meal duration (s), (xiii) median number of visits per meal, (xiv) median feed intake per meal (kg), (xv) median percentage of time spent feeding per meal, calculated as the proportion of the meal's total duration during which the cow was actively occupying a feed bin (total visit time ÷ meal duration × 100%), and (xvi) median number of non-nutritive visits per meal.

In total, 16 behavioral variables were included in the statistical analyses described in Section 2.4: six basic daily feeding and drinking variables (i–vi), two daily agonistic variables (vii–viii), one non-nutritive visit variable (ix), one feed availability variable (x), and six meal-level variables (xi–xvi). These variables were selected because they capture biologically meaningful aspects of individual feeding, drinking, exploration and social strategies. We recognize that this set is not exhaustive as other variables could be derived from the same data streams. The 'moo4feed' package and the analytical framework described in Section 2.4 are designed to accommodate additional variables, and we encourage future researchers to extend this set and evaluate which further metrics may be informative for individual profiling.

## 2.4 Statistical Analysis

We used Bayesian mixed-effects models to quantify repeatability and predictability. All models were fitted using the brms R package [@burkner2017brms], which implements Bayesian regression models via the Stan programming language.

**Variable distribution assessment.** Before fitting any models, we examined the distribution of each behavioral variable by plotting histograms on both the raw and log-transformed scale. For each variable, we computed skewness, the proportion of zero values, and conducted Shapiro-Wilk normality test. This informed the likelihood family selected for each Bayesian model: variables with approximately symmetric distributions were modeled using a Gaussian likelihood (feed intake, feeding duration, water intake, median percentage of time spent feeding per meal, median proportion of feed remaining when cow starts feeding), while strongly right-skewed variables were modeled using a lognormal likelihood (number of feeding visits, drinking duration, number of drinking visits, total meals, median meal duration, median visits per meal, median feed intake per meal, number of non-nutritive visits, median non-nutritive visits per meal, total actor events, and total reactor events). Three variables modeled with a lognormal likelihood (median non-nutritive visits per meal, total actor events, and total reactor events) were shifted by adding a constant of 1 prior to model fitting, given that the lognormal distribution requires strictly positive values and each of these variables contained zero values. Posterior estimates for these variables were shifted back to the original scale for interpretation.

**Repeatability.** For each of the 16 behavioral variables, a Bayesian mixed model was fitted:

$$
\text{response} \sim \text{DIM} + \text{parity} + \text{THI}_{\text{mean}} + \text{poly}(\text{month},\, 2) + (1 \mid \text{cow})
$$

where DIM (days in milk), parity, and mean daily temperature-humidity index (THI) were fixed effects controlling for changes in physiological state and thermal environment, a second-degree polynomial of calendar month captured seasonal trends, and cow identity was a random intercept. Default brms priors were used. MCMC sampling used 4 chains with 1,000 warmup iterations; total iterations ranged from 6,000 to 17,000 per variable to ensure convergence. Among-individual variation was quantified using two metrics derived from posterior draws: repeatability **(R)** and the coefficient of individual variation **(CVi)** [@hertel2020]. R is the proportion of total variance attributable to between-cow differences:

$$
R = \frac{\text{var.cow}}{\text{var.cow} + \text{var.res}}
$$

where $\text{var.cow}$ is the posterior variance of the cow-level random intercept and $\text{var.res}$ is the residual variance. R is thus confounded by the within-individual variance. A high value of R could mean cows differ greatly from each other, or simply mean that each cow is very consistent within herself. To resolve this ambiguity, we also computed CVi, which calculates among-individual spread relative to the population mean, independently of within-individual consistency:

$$
\text{CVi} = \frac{\sqrt{\text{var.cow}}}{\bar{y}}
$$

where $\bar{y}$ is the mean of the response variable. For lognormal models, CVi was instead computed as

$$
\sqrt{\exp(\text{var.cow}) - 1}
$$

to place the estimate on the original data scale. CVi is interpretable across behaviors with different units and facilitates comparison across studies. Model convergence was assessed via R-hat (< 1.01), Bulk- and Tail-Effective Sample Size (ESS), trace plots, and posterior predictive checks [@vehtari2021rank; @gabry2019visualization]. Individual-level posterior distributions were visualized using ridge plots ordered by each cow's posterior mean [@hertel2020].

**Predictability.** Within-individual variance was quantified using a double hierarchical generalized linear model (DHGLM) implemented in brms. In addition to modeling the mean response, brms allows a variance structure to also be imposed on the residual variance (i.e., within-individual variance). Each DHGLM extended the repeatability model by adding a cow-level random effect to the residual variance, partitioning it per individual so that cows with high residual variance are identified as more unpredictable than cows with low residual variance:

$$
\text{sigma} \sim (1 \mid \text{cow})
$$

Population-level predictability was summarized using two metrics derived from the posterior SD of the cow-level random effect in the sigma sub-model ($\text{sd.sigma.cow}$). The residual intra-individual variability **(rIIV)** captures how much cows differ from one another in their within-individual consistency [@hertel2020]:

$$
\text{rIIV} = \exp(\text{sd.sigma.cow})^2
$$

We also computed the coefficient of variation of predictability **(CVP)**, which can be used to compare predictability across different behavioral traits and studies, independent of the scale of the trait:

$$
\text{CVP} = \sqrt{\exp(\text{sd.sigma.cow}^2) - 1}
$$

Both rIIV and CVP are monotonic transformations of the same parameter ($\text{sd.sigma.cow}$). We report both because rIIV has a direct biological interpretation as a fold-ratio in residual SD across individuals, whereas CVP provides a standardized, scale-free quantity for comparison across traits and studies. MCMC sampling settings were identical to those used for the repeatability models.

A high CVP value means that cows differ considerably from each other in how consistent their day-to-day behavior is: for example, some cows may eat similar amounts of feed every day, while others fluctuate their intake substantially across days. A low CVP value means that cows are similar to each other in their predictability: whether that means all cows are highly consistent, or all cows are highly variable is not known, but there is little difference between individuals in their within-individual consistency.

**Behavioral variable clustering.** To organize the 16 behavioral variables by how informative they are for individual profiling, we applied k-means clustering (k = 3, 50 random starts) to the posterior mean CVi and R values for each variable, after scaling both dimensions (@fig-rep-cvp-scatter). All variables had low predictability values, thus measures of predictability were not used in behavioral variable clustering. The number of clusters was selected using the silhouette method [@rousseeuw1987silhouettes], evaluated over k = 2 to 10. The mean silhouette width was highest at k = 3 and k = 5; we chose k = 3 as the more parsimonious solution given the small number of variables (n = 16).

**Individual-level behavioral profiles.** For each cow and each behavioral variable in the cluster with the highest CVi and R (i.e., Cluster 1), we extracted the individual intercept from the posterior distribution: the population intercept plus each cow's random effect (back-transformed to the original scale for lognormal models). This provided us with a representation of a cow's average behavioral type (i.e., a cow's typical expression of their behavior). For example, for daily feed intake modeled with a Gaussian likelihood, if the population intercept is 65 kg and cow A's random effect is 5 kg, cow A's individual intercept would be 70 kg, representing her typical daily feed intake after accounting for DIM, parity, THI, and seasonal effects.

**PCA and cow clustering using k-means.** To synthesize the multi-dimensional behavioral profiles into a small number of interpretable dimensions and identify distinct cow types, we applied principal component analysis (PCA) with varimax rotation [@budaev2010using] to the matrix of individual intercepts for the Cluster 1 variables (one row per cow, one column per variable). For example, if Cluster 1 contained 8 variables and the study retained 107 cows, the PCA input would be a 107 × 8 matrix where each entry is the individual intercept (extracted as described above) for one cow on one behavioral variable. The number of components retained was chosen to explain at least 80% of the cumulative variance [@cangelosi2007component].

Following the PCA, cows were grouped into behavioral types by applying k-means clustering to the retained rotated component scores. The optimal number of clusters $k$ was determined using the silhouette method [@rousseeuw1987silhouettes], evaluated across 100 independent runs with different random seeds, each using 100 random centroid initializations for stability. For each run, the value of $k$ yielding the highest mean silhouette score was selected, and the $k$ chosen most frequently across all seeds was used for the final clustering. The mean silhouette score $\bar{s}$ across all cows serves as an overall measure of cluster quality, reflecting both within-cluster cohesion and between-cluster separation. The resulting clusters were visualized using component score biplots and an interactive 3D scatter plot for exploration of cow clusters across the principal components.

## 2.5 Workflow and Software

To support transparency and reproducibility, all data processing, variable extraction, and analytical steps were implemented in the moo4feed R package <https://www.skysheng.io/moo4feed/>. The functions implemented in this R package have been tested with 96% code coverage (code coverage report: <https://app.codecov.io/gh/skysheng7/moo4feed>). This R package enables researchers to replicate our variable extraction pipeline on their own datasets and to extend the approach to additional behavioral measures. The source code and example data included in the R package is hosted on GitHub: <https://github.com/skysheng7/moo4feed.git>. We used the newly developed R package to extract the behavioral variables described in Section 2.3 and conducted all statistical analyses described in Section 2.4. Data and code for all the data analysis are available at: <https://github.com/skysheng7/competition_dominance_analysis.git>. We have also used Borealis to archive the R package at: <https://doi.org/10.5683/SP3/LF4DRO> and data analysis code at: <https://doi.org/10.5683/SP3/K6J350>.
